Region-network sparsity priors for\\ high-dimensional inference in fMRI data

### 2015-12-03 paper/paper.tex
interm

\textbf{\\keywords}:
Brain Imaging, Functional Specialization, Functional Integration,
Sparsity-inducing Norms, Numerical Optimization, Systems Neuroscience

@@

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions probably solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is probably enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of \textit{structure} (i.e., histological preparation),
\textit{connectivity}, (i.e., axonal tracing),
and \textit{functional properties}
(i.e., single cell recordings) in same animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local cyto- and chemoarchitectonic infrastructure
and its unique global connectivity profile
together realize computational processes (i.e., function).
While
cortical modules versus connections between those
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation},
both architectural principles are conceptually inextricable when
explaining the emergence of mental operations
\cite{tononi1998complexity}.

@@

@@

neurobiologically motivated restrictions to complexity circumvented
the curse of dimensionality

\paragraph{Numerical Optimization}

@@

@@

@@

\begin{equation}
  \argmin 0.5||u-v||_2^2 + \lambda\sum_{g \in T} \eta_g||v_g||_2
\end{equation}

@@

@@

@@

>>>>

\textbf{\\keywords}:
Brain Imaging, Functional Specialization, Functional Integration,
Sparsity-inducing Norms, Structured Sparsity,
Numerical Optimization, Systems Neuroscience

@@

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions probably solve distinct classes of computational processes
\cite{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is probably enabled by complex connections between these
distinct brain regions \cite{sporns14nn}.
%
These notions were predominantly derived from
invasive examination of \textit{structure} (i.e., histological preparation),
\textit{connectivity}, (i.e., axonal tracing),
and \textit{functional properties}
(i.e., single cell recordings) in same animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\cite{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\cite{passingham2002, young93monkey, scannell95cat}.
%
Both
local cyto- and chemoarchitectonic infrastructure
and its unique global connectivity profile
together realize computational processes (i.e., function).
While
cortical modules versus connections between those
reflect 
functional specialization versus functional integration
\cite{friston2002beyond, mesulam_sensation},
both architectural principles are conceptually inextricable when
explaining the emergence of mental operations
\cite{tononi1998complexity, saygin2012}.

@@

higher-order prior knowledge

structured sparsity
extend trend in stastical modelling from
genetics \cite{rapaport2008classification, kim2012tree},
natural image processing \cite{harzallah2009combining}

to brain imaging domain.

@@

neurobiologically motivated restrictions to complexity circumvented
the curse of dimensionality
three-dimensional spatial arrangement that respects
the functional anatomy of the brain
not ignore the spatial configuration

incorporate rich prior knowledge

If meaningful structures exist,
ÇÇçwe show that one can take advantage of such structures

\paragraph{Problem formulation}

Sparse linear models
encode geometric prior information
topology
local sets of voxels

structured regularization
We might therefore gain in the quality of the factors
induced by enforcing directly this a priori

l1 = unstructured sparsity-inducing penalty

@@

hierarchial tree = more generally into a directed acyclic graph

@@

The scalar b is not particularly informative,
however the vector w corresponds to a volume that
can be represented in brain space as a volume
for visualization of the predictive pattern of voxels.

each row of X is a p-dimensional sample,
i.e., an activation map of p voxels related to one stimulus presentation.

@@

The essential shortcoming of the Elastic net is that
it does not take into account the spatial structure of the data,
which is crucial in this context

Craddock clusters are often used for feature agglomeration
into parcels
-> exploits only a part of the data

dual-level spatial structure
sparse hierarchical regularization
structured sparsity-inducing regularization
the root of the tree T is the unique cluster that gathers all the voxels,

It is a generalization of the traditional $\ell_1$-norm
$\Omega(\mathbf{w}) = \sum_{j=1}^p|\mathbf{w}_j|$
ignores structure

\cite{jenatton2011multi}

\textit{structured sparsity}

\cite{huang2011learning, morales2010family, jenatton2011structured}

a node $j$ of $\mathcal{T}$,
we denote by $g_j \subseteq \{1,..,q\}$ the set of indices that record
all the descendants of $j$ in $\mathcal{T}$

the family of sparsity-inducing norms has recently been extended
by hierarchical sparsity penalty terms
\cite{zhao2009composite}.

\begin{equation}
  \Omega(\mathbf{w}) = \sum_{g \in G}||\mathbf{w}_g||_2 = \sum_{g \in G}\sqrt{\sum_{j \in g}\mathbf{w}_j^2}
\end{equation}

For example, when G is the set of all singletons, Ω is the
usual l1 norm (assuming that all the weights are equal to 1).

l1/l2 mixed norm is convex

Discarding coefficients belonging to a network group will naturally enforce
discarding the coefficients belonging to each of its descendent region groups.
Conversely,
variable selection of a network group will also enforce
selection of all voxel of its descendent group regions.
Single region groups can however be set to zero (unselected)
or non-zero (selected)
without analogous effect on the parent network group.

At the between-group level,...
$\Omega$ exerts $\ell_1$-like variable selection on
the (||\mathbf{w}_g||_2)_{g\inG} groups,
yielding a maximum of $g \in G$ to be zeroed out
\cite{jenatton2011structured}.
The important consequence is that also all descentes of such a zeroed
group $g \in G$ will be descarded.
Conversely,
if one group $g$ is selected,
then all the ancestral groups will also be selected.
Thus, statistical estimation will be improved by enticing
entire voxel sets to be selected or discarded as predictive, 
although one individual coefficient is computed for each voxel.

-> it is a $(\ell_1, \ell_2)$-mixed norm
-> between-group sparsity effect by l1
-> within-group shrinkage effect by l2

\begin{equation}
  \Omega(\mathbf{w}) = \sum_{g \in G} \eta_g ||\mathbf{w}_g||_2
\end{equation}

(\eta_g)_{g \in G} are positive weights for the groups

fit to the data is measured through
a convex loss function (w, b) 􏰀→ L(y, X, w, b) ∈ R+.

\paragraph{Classification}

logistic loss function

\begin{equation}
  P(y=k|x, W, b) = \frac{exp\{x^Tw^k + b_k\}}{\sum_{m=1}^cexp\{x^Tw^m + b_m\}}

  \argmin \frac{1}{2}||u-v||_2^2 + \lambda\Omega(\mathbf{w})

\end{equation}

\lambda > 0.

bias is omitted because \mathbf{X} and \mathbf{y}
are mean-centered and unit-variance scaled.

In this setting, and given a new fMRI volume x,
we make predictions by choosing the label that maximizes
the class-conditional probabil- ities (3.1), that is, argmaxk∈{1,...,c}Prob(y = k|x; W∗, b∗)

One-versus-rest scheme

\paragraph{Regression}

squared error as loss

\begin{equation}
  \argmin \frac{1}{2}||\mathbf{y - Xw}||_2^2 + \lambda\Omega(\mathbf{w})
\end{equation}

\lambda > 0.

Prediction for a new fMRI volume x is then simply performed by computing
the dot product x⊤w∗

\paragraph{Numerical optimization}

Difficult because high-dimensional setting

empirical risk minimization was performed by

The intercept $b$ is left unregularized

@@

exploit known structure for variable selection and prediction
in classification and regression

@@

alternatives would be dimensionality reduction
and dictionary learning

@@

More generally, our regularization scheme could also be used for various
learning tasks, as soon
as prior knowledge on the structure of the sparse representation
is available, e.g.,
for multiple kernel learning (Micchelli and Pontil, 2006),
multi-task learning (Argyriou et al., 2008; Obozinski et al., 2009;
Kim and Xing, 2009) and sparse matrix factorization
problems (Mairal et al., 2010; Jenatton et al., 2010).

this type of prior knowledge is of clear interest for the spatially and tempo-
rally structured data typical in bioinformatics, computer vision and neuroscience applications

structured sparsity can even be used for
unsupervised data modelling,
such as in PCA \cite{jenatton2009structured}
